---
title: "Comparison of Field vs. Lab results -- pH and EC25"
output: 
  html_notebook:
    df_print: paged
    toc_depth: 4
---

## Step 0: Import Data

```{r Import1, eval=FALSE}
library(openxlsx)
# Note: This file will _not_ be in the Project Folder, but the loaded DF _will be_
filename <- choose.files(caption="Select xlsx File to Import", multi=FALSE)
DATA.file <- read.xlsx(filename, sheet="Results", startRow = 1, colNames=TRUE, rowNames=FALSE, detectDates=FALSE);  
```

Working datafile has `r format(nrow(DATA.file), big.mark=",")` *rows* and `r length(names(DATA.file))` *columns*.

```{r DATAFILE-1, eval=FALSE}
file0 <- regexpr("[.]",basename(filename))
file1 <- substr(basename(filename),1,max(file0[[1]])-1)
PATH.file1 <- dirname(filename)
```

### Examine Dataset Characteristics

```{r DATAFILE-char, eval=FALSE}
data.Vars <- data.frame()
for (i in 1:length(names(DATA.file))) {
  class_i <- class(DATA.file[,i])
  name_i <- names(DATA.file[i])
  num.NAs <- sum(is.na(DATA.file[,i]))
  count_levels <- length(unique(DATA.file[,i]))
  num.blanks <- length(DATA.file[DATA.file[,i] == "",c(i)])
data.Vars <- rbind(data.Vars, data.frame(i, name_i, class_i, num.NAs, count_levels,num.blanks)) }
print(data.Vars)
# Save variable summary (DF) to a File
#filename_1 <- paste(file1, "_VarLEVELS.xlsx", sep='')
#write.xlsx(data.Vars, file=filename_1, sheetName="Var.Chars")
```

### Identify Observations with paired Field / Lab measurements of a parameter, by day

**Key variables:**  

* Monitoring.Location.ID
* Monitoring.Location.Type
* Activity.Type
* Activity.Start.Date
* Characteristic.Name
* Detection.Condition
* Result.Qualifier
* Result.Value
* Result.Unit
* Detection/Quantitation.Limit.Type1
* Detection/Quantitation.Limit.Value1
* Analytical.Method.ID
* Analytical.Method.Name
  
**Derived variables:**

1. Measure.Type == Field // Lab
2. Site.Type == Riv.Strm // Lake // Other
3. ResultX == Result.Value // MDL // UQL
4. Month %from% DATE

```{r DERIVE, eval=FALSE, message=FALSE}
# (2) from levels of DATA.file$Monitoring.Location.Type
lakes <- c("Lake")
streams <- c("River/Stream", "Canal Drainage", "Canal Irrigation", "Canal Transport", "Storm Sewer", "Waste Sewer")
other <- c("Facility Other", "Mine/Mine Discharge", "Other-Ground Water", "Seep", "Spring",
           "Well", "Wetland Undifferentiated")
#
DATA.file$Site.Type <- "Other"
DATA.file$Site.Type <- ifelse(DATA.file$Monitoring.Location.Type %in% lakes,
                              "Lake",
                              ifelse(DATA.file$Monitoring.Location.Type %in% streams,
                                     "Stream",
                                     "Other"))
# (1) from levels of Activity.Type
library(dplyr)
DATA.file$Measure.Type <- "Other"
DATA.file$Measure.Type <- ifelse((grepl("^Sample-", DATA.file$Activity.Type)),
                                 "Lab",
                            ifelse((grepl("^Field", DATA.file$Activity.Type)), 
                                   "Field",
                                   "Other"))
# (3) from Result value, DetCond and Res.Qualifier
DATA.file$ResultX <- as.numeric(DATA.file$Result.Value)
DATA.file$ResultX <- ifelse(is.na(DATA.file$Result.Value),
                            as.numeric(DATA.file$`Detection/Quantitation.Limit.Value1`),
                            as.numeric(DATA.file$Result.Value))

# (4) from Activity.Start.Date
DATA.file$Date <- as.Date(DATA.file$Activity.Start.Date, origin="1899-12-13")
DATA.file$Year <- as.numeric(format(DATA.file$Date, "%Y"))
DATA.file$Month <- as.numeric(format(DATA.file$Date, "%m"))
DATA.file$YYMM <- DATA.file$Year + DATA.file$Month/100
```

### Trim dataset
```{r DATA.trim, eval=FALSE}
DATA.file2 <- DATA.file[,c(2,4,5,6,21,24,25,26,27,29,30,31,32,34,36,39:45,1)]
# clean up headers
names(DATA.file2)[1] <- "MLID"
names(DATA.file2)[2] <- "MLID.type"
names(DATA.file2)[6] <- "DET.Cond"
names(DATA.file2)[7] <- "RES.qual"
names(DATA.file2)[11] <- "RES.comm"
names(DATA.file2)[12] <- "DETLIM.type"
names(DATA.file2)[13] <- "DETLIM.val"
```

New dataset is: `DATA.file2`, with `r format(nrow(DATA.file2), big.mark=",")` rows and `r length(names(DATA.file2))` columns.

### Generate Observation Key

Combine:  

* MLID
* Activity.Start.Date
* Year & Month (truncated)

```{r DATA.keys, eval=FALSE}
DATA.file2$Samp.key <- paste(DATA.file2$MLID, DATA.file2$Activity.Start.Date, sep=".")
DATA.file2$Samp.key2 <- paste(DATA.file2$MLID, DATA.file2$YYMM, sep=".")
```

## Identify Paired Field-Laboratory Observations

Also need to remove data logger results, samples with MLID = NA, and other cleanup

```{r DATA.fix, eval=FALSE}
DATA.file3 <- DATA.file2[DATA.file2$Activity.Type != "Field Msr/Obs-Portable Data Logger",]
DATA.file3 <- DATA.file3[DATA.file3$Activity.Type != "Quality Control Field Replicate Portable Data Logger",]
DATA.file3 <- DATA.file3[complete.cases(DATA.file3[1]),]
DATA.file3 <- DATA.file3[DATA.file3$Activity.Type != "Sample-Integrated Vertical Profile",]

## examine dataset structure (again)
data.Vars.3 <- data.frame()
for (i in c(1:18,20:25)) {
  class_i <- class(DATA.file3[,i])
  name_i <- names(DATA.file3[i])
  num.NAs <- sum(is.na(DATA.file3[,i]))
  count_levels <- length(unique(DATA.file3[,i]))
  num.blanks <- length(DATA.file[DATA.file3[,i] == "",c(i)])
data.Vars.3 <- rbind(data.Vars.3, data.frame(i, name_i, class_i, num.NAs, count_levels,num.blanks)) }
```

### Separate Data by Parameter (pH / EC25)

```{r DATA.subset0, eval=FALSE}
PH.dat0 <- DATA.file3[DATA.file3$Characteristic.Name == "pH",c(-11, -12, -14, -15)]
PH.dat0 <- droplevels.data.frame(PH.dat0)
#
EC.dat0 <- DATA.file3[DATA.file3$Characteristic.Name == "Specific conductance",
                      c(-11, -12, -14, -15)]
EC.dat0 <- droplevels.data.frame(EC.dat0)

```

## Examine patterns between Field and Lab

### Cross-tab (dcast) results by Measure.Type

```{r DATA.cross, eval=FALSE, message = FALSE}
# pH
library(reshape2)
PH.dat1 <- dcast(PH.dat0, Samp.key + Samp.key2 + Site.Type ~ Measure.Type,
                 value.var="ResultX", fun.aggregate=function(x) mean(x, na.rm=TRUE))
#
EC.dat1 <- dcast(EC.dat0, Samp.key + Samp.key2 + Site.Type ~ Measure.Type,
                 value.var="ResultX", fun.aggregate=function(x) mean(x, na.rm=TRUE))

```

### Check correctness of values:

* pH: values b/w 3.0 and 12.0
* EC: values b/w 50 and 100,000 (+/-)

pH Cross-tab file :: **`PH.dat1`**

```{r DATA.PHvalues}
#summary(PH.dat1)
options(width=95)
library(tidyr)
library(htmlTable)
htmlTable(summary(PH.dat1), caption="pH cross-tab summary", css.cell="padding-left: 0.5em; padding-right: 2em;", rnames=FALSE, align="l")
```

#### pH:: Check out values and trim results to reasonable ranges

```{r DATA.cdf1}
#
library(ggplot2)
library(scales)
ggplot() + theme_bw() + stat_ecdf(data=subset(PH.dat1, !is.na(Field) & Field > 3 & Field < 12),aes(x=Field), na.rm=T, color="red") + stat_ecdf(data=subset(PH.dat1, !is.na(Lab) & Lab > 3 & Lab < 12), aes(x=Lab), na.rm=T, color="green") + scale_x_continuous(limits=c(3,12)) + ggtitle("Field and Lab pH values, with range 2.0 to 12.0")
#
PH.dat1[(PH.dat1$Field < 2 & !is.na(PH.dat1$Field)),c(1,3,4,5)]
PH.dat1[(PH.dat1$Field > 12 & !is.na(PH.dat1$Field)),c(1,3,4,5)]
```

**Plot Lab (y) versus Field (x) pH values**

```{r FIGS1, eval=FALSE}
library(ggplot2)
PH.dat2 <-PH.dat1[(!is.na(PH.dat1$Field) & !is.na(PH.dat1$Lab) & PH.dat1$Field > 2 & PH.dat1$Field < 12 & PH.dat1$Lab > 2 & PH.dat1$Lab < 12),]
```


```{r FIGS2, }
library(tidyr)
library(htmlTable)
htmlTable(summary(PH.dat2[c(4,5)]), caption="Summary of trimmed pH values", css.cell="padding-left: 0.5em; padding-right: 2em;", rnames=FALSE, align="l")
#
ggplot(data = PH.dat2, aes(x=Field, y=Lab, col=Site.Type)) + geom_point() + geom_smooth(method="lm") + geom_abline(slope=1, intercept=0, lty=2, col="black") + theme_bw() + ggtitle("Lab vs. Field pH for Lakes, Streams, and other water types")
```

#### Additional data exploration of pH values can be performed here...

xxx
***

#### EC25::  Check out values and trim results to reasonable ranges

```{r DATA.ECvalues,}
options(width=95)
library(tidyr)
library(htmlTable)
htmlTable(summary(EC.dat1), caption="EC25 cross-tab summary", css.cell="padding-left: 0.5em; padding-right: 2em;", rnames=FALSE, align="l")

```

```{r DATA.ecdf2}
options(scipen = 999)
library(ggplot2)
library(scales)
ggplot() + theme_bw() + stat_ecdf(data=subset(EC.dat1, !is.na(Field) & Field > 0), aes(x=Field), na.rm=T, color="red") + stat_ecdf(data=subset(EC.dat1, !is.na(Lab) & Lab > 0), aes(x=Lab), na.rm=T, color="green") + ggtitle("Field & Lab EC25 values") + scale_x_log10()
```

** Constrain EC data and plot Lab (y) versus Field (x) values**

```{r FIGS3.dat, eval=FALSE}
EC.dat2 <- EC.dat1[(!is.na(EC.dat1$Field) & !is.na(EC.dat1$Lab) & EC.dat1$Field > 10 & EC.dat1$Lab > 10),]
```

```{r FIGS4, }
#
library(tidyr)
library(htmlTable)
htmlTable(summary(EC.dat2[c(4,5)]), caption="Summary of trimmed EC25 values", css.cell="padding-left: 0.5em; padding-right: 2em;", rnames=FALSE, align="l")
#
library(ggplot2)
library(scales)
ggplot() + theme_bw() + stat_ecdf(data=subset(EC.dat2, !is.na(Field) & Field > 10), aes(x=Field), na.rm=T, color="red") + stat_ecdf(data=subset(EC.dat2, !is.na(Lab) & Lab > 10), aes(x=Lab), na.rm=T, color="green") + ggtitle("Field & Lab EC25 values") + scale_x_log10() + coord_trans()
```

```{r EC.ecdf,}
library(scales)
ggplot(data = EC.dat2, aes(x=Field, y=Lab, col=Site.Type)) + geom_point() + geom_smooth(method="lm") + geom_abline(slope=1, intercept=0, lty=2, col="black") + theme_bw() + ggtitle("Lab vs. Field EC25 for Lakes, Streams, and other water types") + scale_x_continuous(labels=comma) + scale_y_continuous(labels=comma)
```



***

